---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
install.packages("viridis")
library(viridis)
install.packages("RColorBrewer")
library(RColorBrewer)
install.packages("ggrepel")
library(ggrepel)
library(scales)

```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
expeditions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv')
glimpse(expeditions)

theme_set(theme_light())

expeditions %>%
  filter(peak_name == fct_lump_n(peak_name, n = 15)) %>%
  group_by(season) %>%
  count(peak_name) %>%
  mutate(peak_desc = fct_reorder(peak_name, n)) %>%
  ggplot(aes(x = reorder(peak_desc, n), y = n, fill = season)) +
  geom_col() +
  coord_flip() +
  labs(title = "The 15 most popular peaks stacked by season of expedition", x = NULL, y = "Number of expeditions") +
  theme(legend.position = "bottom") +
  scale_fill_viridis(discrete = TRUE)
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phd_by_field <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv")
glimpse(phd_by_field)

theme_set(theme_minimal())

phd_df <- phd_by_field[ , c('broad_field', 'year', 'n_phds')]

phd_df %>%
  filter(n_phds != "NA") %>%
  group_by(broad_field, year) %>%
  mutate(year = as.Date(year),
         n_phds = sum(n_phds)) %>%
  ggplot(aes(x = year, y = n_phds, color = broad_field, group = broad_field)) +
  geom_line(linewidth = 1.2) +
  guides(color = guide_legend(title = "Broad field")) +
  scale_x_continuous(breaks = breaks_pretty()) +
  scale_y_continuous(labels = comma) +
  scale_color_brewer(palette = "Dark2") +
  labs(fill = 'Broad field', title = "Number of awarded Ph.D.-s in the US by year",  x = NULL, y = NULL) +
  theme(panel.border = element_blank())
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commute <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv")
glimpse(commute)

theme_set(theme_light())

commute[commute$state=='Ca', 'state'] <- 'California'
table(commute$state)

commute %>%
  group_by(state, mode, state_region, state_abb) %>%
  summarise(sum = sum(n)) %>%
  spread(key = mode, value = sum) %>%
  ggplot(aes(Walk, Bike, color = state_region)) +
  geom_point(size = 2) +
  guides(color = guide_legend(title = "State region")) +
  scale_x_log10(label=comma) +
  scale_y_log10(label=comma) +
  labs(fill = "State region", title = "Number of people walking vs. biking to work in each USA state", x = "Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)") +
  geom_text(aes(label = state_abb), color = "black", show.legend = FALSE, check_overlap = TRUE)

```
